Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics

### 2017-04-02 rigid_bodies.tex
introduction

\title{Revisiting the Nos\'{e}-Hoover chain thermostat applied to the Rigid Bodies Molecular Dynamics}

@@

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that makes possible a microscopic insight into complex physical phenomena. To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid. For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integration of the equations of motion in Molecular Dynamics disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we observed the same problem, but regarding the partition of the kinetic energy between translational and rotational degrees of freedom in Molecular Dynamics of rigid bodies in the microcanonical ensemble. That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body Molecular Dynamics stems from the fact that its application is not limited to molecules actually modelled as a rigid body, like the ubiquitous case of water, but also it constitues a simple coarse-grain strategy for systems that might be designed as collection of rigid bodies interconected, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}. As a matter of fact, the approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has been applied in the simulation of small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as in membranes~\cite{Bucior_2012} molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}, DNA\cite{Fathizadeh_2012}

The Nos\'{e}-Hoover chain thermostat\cite{Martyna1992}, routinely used in Molecular Dynamics, employs the extended phase-space approach and introduces ficticius variables in order to ``control'' the kinetic energy fluctuations of the physical variables which allows to maintain the system at a specified temperature. The dynamics of those variables introduces non-Hamiltonian components which makes infeasible to carry out the backward error analysis. Altough it is not possible to devise symplectic integrator, one can rigurously obtain measure-preserving numerical solvers\cite{Sergi2001,Ezra2004,Ezra2006} and obtain numerical stability.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble and introduce time reversible and measure-preserving numerical integrators as well. Our startint point is the symplectic integrator in the microcanonical ensemble we introduced in a previous paper\citep{Abreu_2017}, which employs the unit quaterion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at the specified value. The analysis of the thermostats will be rather heuristic reliying on comparisons with results obtained employing the Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present. We will also include results from the SHAKE\cite{Ryckaert1977} method, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints. unavoidable errors but we shoud be aware of the effect on the properties of interest.

>>>>

\title{Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics}

@@

Molecular Simulation, being the computational realization of Statistical Mechanics\cite{Tuckerman2010}, involves several types of approximations in order to mimic ``real'' physical systems. Naively, one would expect those approximations to only produce deviation from experimental values. However, they actually may give rise to systematic errors introducing non-physical behaviour. For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics, disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we reported the same artificiality to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics. That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.

The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water, but also it is a simple coarse-grain strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}. As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented in the softwares packages \textit{blue}-HOOMD and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as membranes~\cite{Bucior_2012}, molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}.

The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables. The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with what backward error analysis is not feasible anymore. Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}

In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical ensemble. The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom. Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value. The analysis of the integrators will be rather heuristic, relying on comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy. We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
